Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2021-05-01 ms/Thesis/thesis.tex
Add comment about stable and unstable polymorphism
Thesis: add few of the fixes that we did before publishing to PRSB

The role of assortment in the evolution of altruism was emphasized by \citet{Fletcher2009assortment}.
They found that in a \emph{public-goods} game, altruism will evolve if cooperative individuals experience more cooperation, on average, than defecting individuals, and ``thus, the evolution of altruism requires (positive) assortment between focal \emph{cooperative} players and cooperative acts in their interaction environment.''
With some change in parameters, this condition is summarized by \citep[eq.~2.3]{Fletcher2009assortment}
\begin{equation} \label{eq:fletcher2009}
c < b \cdot (p_C - p_D ) \,,
\end{equation}
where $p_C$ is the probability that a cooperator receives help, and $p_D$ is the probability that a defector receives help.\footnote{Inequality~\ref{eq:fletcher2009} generalizes inequalities~\ref{eq:hamilton_rule} and~\ref{eq:eshel1982} by substituting $p_C=r + p$, $p_D=p$ and $p_C=m + (1-m)p$, $p_D=(1-m)p$, respectively, where $p$ is the frequency of cooperators.}
\citet{Bijma2010assortment} obtained a result related to inequality~\ref{eq:fletcher2009} for other types of games.

\citet{feldman1985gene} introduced the first model for the evolution of altruism by cultural transmission with kin selection and demonstrated that if the fidelity of cultural transmission of altruism is $\varphi$, then the condition for evolution of altruism in the case of sib-to-sib altruism is \citep[Eq.~16]{feldman1985gene}
\begin{equation} \label{eq:feldman1985}
c < b \cdot \varphi - \frac{1-\varphi}{\varphi} \,.
\end{equation}
In inequality~\ref{eq:feldman1985}, $\varphi$ takes the role of relatedness ($r$ in inequality~\ref{eq:hamilton_rule}) or assortment ($m$ in inequality~\ref{eq:eshel1982}), but the effective benefit $b\cdot \varphi$ is  reduced by $(1-\varphi)/\varphi$.
This shows that under a cultural transmission, the condition for the evolutionary success of altruism entails a modification of Hamilton's rule (inequality~\ref{eq:hamilton_rule}).

@@

Social interactions occur randomly:
two juvenile individuals with phenotype $A$ interact with probability~$\hat{p}^2$, two juveniles with phenotype $B$ interact with probability $(1-\hat{p})^2$, and two juveniles with different phenotypes interact with probability $2\hat{p}(1-\hat{p})$. 
Horizontal cultural transmission occurs between pairs of individuals from the same generation. 
It occurs between socially interacting partners with probability~$\alpha$, or between a random pair with probability~$1-\alpha$ (see~\autoref{fig:illustration}b).
%The interaction-transmission association $\alpha$ is therefore the fraction of population that receives (horizontal transmission) from the social interaction partner, and $1-\alpha$ receives randomly.
However, horizontal transmission is not always successful, as one partner may reject the other's phenotype.
The probability of successful horizontal transmission of phenotypes $A$ and $B$ are $T_A$ and $T_B$, respectively (\autoref{table:interactions}, \autoref{fig:illustration}c).
Then, the frequency $p'$ of phenotype $A$ among adults in the next generation, after horizontal transmission, is 
\begin{equation}\label{eq:nextgen_adults}
\begin{aligned}
p' \;= \;
& \hat{p}^2 \big[\alpha + (1-\alpha)\big(\hat{p} + (1-\hat{p})(1-T_B)\big)\big] \;+ \\
& \hat{p}(1-\hat{p}) \big[\alpha(1-T_B) + (1-\alpha)\big(\hat{p} + (1-\hat{p})(1-T_B)\big)\big] \;+ \\
& (1-\hat{p})\hat{p} \big[\alpha T_A + (1-\alpha) \hat{p} T_A \big] \;+ 
(1-\hat{p})^2 \big[(1-\alpha) \hat{p} T_A \big] \\
=\; & \hat{p}^2(T_B-T_A) \;+\; \hat{p}(1+T_A-T_B) \;.
\end{aligned}
\end{equation}
The frequency of $A$ among parents (i.e. after selection) follows a similar dynamic, but also includes the effect of natural selection, and is therefore
\begin{equation}\label{eq:nextgen_parents}
\begin{aligned}
\bar{w} \dot{p}' =
& \hat{p}^2 (1+b-c) \big[\alpha + (1-\alpha)\big(\hat{p} + (1-\hat{p})(1-T_B)\big)\big] \;+ \\
& \hat{p}(1-\hat{p}) (1-c) \big[\alpha(1-T_B) + (1-\alpha)\big(\hat{p} + (1-\hat{p})(1-T_B)\big)\big] \;+ \\
& (1-\hat{p})\hat{p} (1+b) \big[\alpha T_A + (1-\alpha) \hat{p} T_A \big] \;+\; 
(1-\hat{p})^2 \big[(1-\alpha) \hat{p} T_A \big] \;,
\end{aligned}
\end{equation}
where fitness values are taken from \autoref{fig:illustration}a and \autoref{table:interactions}, and the population mean fitness is
$\bar{w} =  1 + \hat{p}(b-c)$.
Starting from \autoref{eq:horizontal} with $\hat{p}'=v\dot{p}'+(1-v)p'$, we substitute $p'$ from \autoref{eq:nextgen_adults} and $\dot{p}'$ from \autoref{eq:nextgen_parents} and obtain
\begin{equation} \label{eq:nextgen_juveniles}
\begin{aligned}
\hat{p}' =
& \frac{v}{\bar{w}}\Big[\hat{p}^2(1+b-c)\Big(1-(1-\hat{p})(1-\alpha)T_B)\Big)\Big] \;+ \\
& \frac{v}{\bar{w}}\Big[ \hat{p}(1-\hat{p})(1-c)\big( \hat{p}(1-\alpha)T_B + 1 - T_B \big) \Big] \;+ \\
& \frac{v}{\bar{w}}\Big[ \hat{p}(1-\hat{p})(1+b)\big(\hat{p}(1-\alpha) + \alpha \big) T_A \Big] \;+ \\
& \frac{v}{\bar{w}}(1-\hat{p})^2\hat{p}(1-\alpha)T_A \;+\;
(1-v)\hat{p}^2(T_B-T_A) \;+\;
(1-v)\hat{p}(1+T_A-T_B) \;.
\end{aligned}
\end{equation}
\autoref{table:vars_params} lists the model variables and parameters.

@@

\end{result}
These conditions are illustrated in Figures~\ref{fig:equilibria}a, \ref{fig:equilibria}b, \ref{fig:equilibria_v1}a, and \ref{fig:equilibria_v1}b, and the analysis is in Appendix~\ref{sec:appendixB}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Population structure}
All the simulations in this section were made by Ohad Lewin-Epstein from Tel Aviv University.
Interaction-transmission association may also emerge from population structure.
Consider a  population colonizing a two-dimensional grid of size 100-by-100, where each site is inhabited by one individual, similarly to the model of \citet{lewin2020rockpaperscissors}.
Each individual is characterized by its phenotype: either cooperator, $A$, or defector, $B$.
Initially, each site in the grid is randomly colonized by either a cooperator or a defector, with equal probability.
In each generation, half of the individuals are randomly chosen to "initiate" interactions, and these
initiators interact with a random neighbor (i.e. individual in a neighboring site) in a prisoners' dilemma game (\autoref{fig:illustration}a) and a random neighbor (with replacement) for horizontal cultural transmission (\autoref{fig:illustration}b).
The expected number of each of these interactions per individual per generation is one.
The effective interaction-transmission association $\alpha$ in this model is the probability that the same neighbor is picked for both interactions, or $\alpha=1/M$, where $M$ is the number of neighbors.
On an infinite grid, $M=8$, but on a finite grid $M$ can be lower in edge neighborhoods close to the grid border.
As before, $T_A$ and $T_B$ are the probabilities of successful horizontal transmission of phenotypes $A$ and $B$, respectively.

@@

To understand the role of horizontal transmission, we first review the role of \emph{assortment}.
\citet{Eshel1982} showed that altruism can evolve when the tendency for \emph{assortative meeting}, i.e., for individuals to interact with others of their own phenotype, is strong enough.
\citet{Fletcher2009assortment}  further argued that a general explanation for the evolution of altruism is given by \emph{assortment}: the correlation between individuals that carry an altruistic trait and the amount of altruistic behavior in their interaction group (see also \citet{Bijma2010assortment}).
They suggested that to explain the evolution of altruism, we should seek mechanisms that generate  assortment, such as population structure, repeated interactions, and individual recognition.
Our results highlight another mechanism for generating assortment: an association between social interactions and horizontal transmission that creates a correlation between one's partner for interaction and the partner for transmission.
This mechanism does not require repeated interactions, population structure, or individual recognition.
We show that high levels of such interaction-transmission association greatly increase the potential for evolution of cooperation.
With enough interaction-transmission association, cooperation can increase in frequency when initially rare even when there is horizontal transmission bias against it ($T_A<T_B$).

@@

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Bibliography
\bibliographystyle{rsnat}
\bibliography{bib}

>>>>

The role of assortment in the evolution of altruism was emphasized by \citet{Fletcher2009assortment}.
They found that in a \emph{public-goods} game, altruism will evolve if cooperative individuals experience more cooperation, on average, than defecting individuals, and ``thus, the evolution of altruism requires (positive) assortment between focal \emph{cooperative} players and cooperative acts in their interaction environment.''
With some change in parameters, this condition is summarized by \citep[eq.~2.3]{Fletcher2009assortment}
\begin{equation} \label{eq:fletcher2009}
c < b \cdot (p_C - p_D ) \,,
\end{equation}
where $p_C$ is the probability that a cooperator receives help, and $p_D$ is the probability that a defector receives help\footnote{Inequality~\ref{eq:fletcher2009} generalizes inequalities~\ref{eq:hamilton_rule} and~\ref{eq:eshel1982} by substituting $p_C=r + p$, $p_D=p$ and $p_C=m + (1-m)p$, $p_D=(1-m)p$, respectively, where $p$ is the frequency of cooperators.}.
\citet{Bijma2010assortment} obtained a result related to inequality~\ref{eq:fletcher2009} for other games.

Cooperation can also evolve when interactions are determined by population structure. For example, \citet{Ohtsuki2006} studied populations on graphs with average degree $k$, that is, the average individual has $k$ potential interaction partners. Assuming that selection is weak and that the population size is much larger than $k$ (i.e. sparse structure), they found that cooperative behaviour can evolve if~\citep{Ohtsuki2006}
\begin{equation} \label{eq:ohtsuki2006}
c < b \cdot \frac{1}{k} \,.
\end{equation}
They thus interpret $1/k$ as \emph{social relatedness} or \emph{social viscosity}~\citep{Ohtsuki2006}.

\citet{feldman1985gene} introduced the first model for the evolution of altruism by cultural transmission with kin selection and demonstrated that if the fidelity of cultural transmission of altruism is $\varphi$, then the condition for evolution of altruism in the case of sib-to-sib altruism is \citep[Eq.~16]{feldman1985gene}
\begin{equation} \label{eq:feldman1985}
c < b \cdot \varphi - \frac{1-\varphi}{\varphi} \,.
\end{equation}
In inequality~\ref{eq:feldman1985}, $\varphi$ replaces relatedness ($r$ in inequality~\ref{eq:hamilton_rule}) or assortment ($m$ in inequality~\ref{eq:eshel1982}), but the effective benefit $b\cdot \varphi$ is  reduced by $(1-\varphi)/\varphi$.
This shows that under a cultural transmission, the condition for the evolutionary success of altruism entails a modification of Hamilton's rule (inequality~\ref{eq:hamilton_rule}).

@@

Social interactions occur randomly:
two juvenile individuals with phenotype $A$ interact with probability~$\hat{p}^2$, two juveniles with phenotype $B$ interact with probability $(1-\hat{p})^2$, and two juveniles with different phenotypes interact with probability $2\hat{p}(1-\hat{p})$. 
Horizontal cultural transmission occurs between pairs of individuals from the same generation. 
It occurs between socially interacting partners with probability~$\alpha$, or between a random pair with probability~$1-\alpha$ (see~\autoref{fig:illustration}b).
%The interaction-transmission association $\alpha$ is therefore the fraction of population that receives (horizontal transmission) from the social interaction partner, and $1-\alpha$ receives randomly.
However, horizontal transmission is not always successful, as one partner may reject the other's phenotype.
The probability of successful horizontal transmission of phenotypes $A$ and $B$ are $T_A$ and $T_B$, respectively (\autoref{table:interactions}, \autoref{fig:illustration}c).
Thus, the frequency $p'$ of phenotype $A$ among adults in the next generation, after horizontal transmission, is 
\begin{equation}\label{eq:nextgen_adults}
\begin{aligned}
p' \;= \;
& \hat{p}^2 \big[\alpha + (1-\alpha)\big(\hat{p} + (1-\hat{p})(1-T_B)\big)\big] \;+ \\
& \hat{p}(1-\hat{p}) \big[\alpha(1-T_B) + (1-\alpha)\big(\hat{p} + (1-\hat{p})(1-T_B)\big)\big] \;+ \\
& (1-\hat{p})\hat{p} \big[\alpha T_A + (1-\alpha) \hat{p} T_A \big] \;+ 
(1-\hat{p})^2 \big[(1-\alpha) \hat{p} T_A \big] \\
=\; & \hat{p}^2(T_B-T_A) \;+\; \hat{p}(1+T_A-T_B) \;.
\end{aligned}
\end{equation}
The frequency of $A$ among parents (i.e. after selection) follows a similar dynamic, but also includes the effect of natural selection, and is therefore
\begin{equation}\label{eq:nextgen_parents}
\begin{aligned}
\bar{w} \dot{p}' =
& \hat{p}^2 (1+b-c) \big[\alpha + (1-\alpha)\big(\hat{p} + (1-\hat{p})(1-T_B)\big)\big] \;+ \\
& \hat{p}(1-\hat{p}) (1-c) \big[\alpha(1-T_B) + (1-\alpha)\big(\hat{p} + (1-\hat{p})(1-T_B)\big)\big] \;+ \\
& (1-\hat{p})\hat{p} (1+b) \big[\alpha T_A + (1-\alpha) \hat{p} T_A \big] \;+\; 
(1-\hat{p})^2 \big[(1-\alpha) \hat{p} T_A \big] \;,
\end{aligned}
\end{equation}
where fitness values are taken from \autoref{fig:illustration}a and \autoref{table:interactions}, and the population mean fitness is
$\bar{w} =  1 + \hat{p}(b-c)$.
Starting from \autoref{eq:horizontal} with $\hat{p}'=v\dot{p}'+(1-v)p'$, we substitute $p'$ from \autoref{eq:nextgen_adults} and $\dot{p}'$ from \autoref{eq:nextgen_parents} and obtain
\begin{equation} \label{eq:nextgen_juveniles}
\begin{aligned}
\hat{p}' =
& \frac{v}{\bar{w}}\Big[\hat{p}^2(1+b-c)\Big(1-(1-\hat{p})(1-\alpha)T_B)\Big)\Big] \;+ \\
& \frac{v}{\bar{w}}\Big[ \hat{p}(1-\hat{p})(1-c)\big( \hat{p}(1-\alpha)T_B + 1 - T_B \big) \Big] \;+ \\
& \frac{v}{\bar{w}}\Big[ \hat{p}(1-\hat{p})(1+b)\big(\hat{p}(1-\alpha) + \alpha \big) T_A \Big] \;+ \\
& \frac{v}{\bar{w}}(1-\hat{p})^2\hat{p}(1-\alpha)T_A \;+\;
(1-v)\hat{p}^2(T_B-T_A) \;+\;
(1-v)\hat{p}(1+T_A-T_B) \;.
\end{aligned}
\end{equation}
\autoref{table:vars_params} lists the model variables and parameters.

@@

\end{result}
Thus, cooperation can take over the population if it has either a horizontal transmission advantage, or if it has a horizontal transmission disadvantage, but the vertical transmission rate is high enough.
In either case, the cost of cooperation must be small enough.
A stable polymorphism can exist between cooperation and defection only if defection has a horizontal transmission advantage.
In this case, the existence of a stable polymorphism depends on the interplay between the benefit and cost of cooperation and the vertical transmission rate.
These conditions are illustrated in Figures~\ref{fig:equilibria}a, \ref{fig:equilibria}b, \ref{fig:equilibria_v1}a, and \ref{fig:equilibria_v1}b, and the analysis is in Appendix~\ref{sec:appendixB}.
Note that \emph{stable} and {unstable polymorphism} are also called, respectively, \emph{coexistence} and \emph{bistable competition}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Population structure}
All the simulations in this section were made by Ohad Lewin-Epstein from Tel Aviv University.
Interaction-transmission association may also emerge from population structure.
Consider a  population colonizing a two-dimensional grid of size 100-by-100, where each site is inhabited by one individual, similarly to the model of \citet{lewin2020rockpaperscissors}.
Each individual is characterized by its phenotype: either cooperator, $A$, or defector, $B$.
Initially, each site in the grid is randomly colonized by either a cooperator or a defector, with equal probability.
In each generation, half of the individuals are randomly chosen to "initiate" interactions, and these
initiators interact with a random neighbor (i.e. individual in a neighboring site) in a prisoners' dilemma game (\autoref{fig:illustration}a) and a random neighbor (with replacement) for horizontal cultural transmission (\autoref{fig:illustration}b).
The expected number of each of these interactions per individual per generation is one, but the realized number of interactions can be zero, one, or even more than one, and in every interaction both individuals are affected, not just the initiator.
The effective interaction-transmission association $\alpha$ in this model is the probability that the same neighbor is picked for both interactions, or $\alpha=1/M$, where $M$ is the number of neighbors.
On an infinite grid, $M=8$ (i.e. Moore neighbourhood~\citep{moore1962machine}), but on a finite grid $M$ can be lower in neighbourhoods close to the grid border.
As before, $T_A$ and $T_B$ are the probabilities of successful horizontal transmission of phenotypes $A$ and $B$, respectively.

@@

To understand the role of horizontal transmission, we first review the role of \emph{assortment}.
\citet{Eshel1982} showed that altruism can evolve when the tendency for \emph{assortative meeting}, i.e. for individuals to interact with others of their own phenotype, is strong enough.
\citet{Fletcher2009assortment}  further argued that a general explanation for the evolution of altruism is given by \emph{assortment}: the correlation between individuals that carry an altruistic trait and the amount of altruistic behavior in their interaction group (see also \citet{Bijma2010assortment}).
They suggested that to explain the evolution of altruism, we should seek mechanisms that generate  assortment, such as population structure, repeated interactions, and individual recognition.
Our results highlight another mechanism for generating assortment: an association between social interactions and horizontal transmission that creates a correlation between one's partner for interaction and the partner for transmission.
This mechanism does not require repeated interactions, population structure, or individual recognition.
We show that high levels of such interaction-transmission association greatly increase the potential for evolution of cooperation.
With enough interaction-transmission association, cooperation can increase in frequency when initially rare even when there is horizontal transmission bias against it ($T_A<T_B$).

@@

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Bibliography
\bibliographystyle{unsrtnat}
\bibliography{bib}
